---
title: MCMC diagnostics
output:
  html_document:
    toc: true
    toc_float: true
---

# Goals 

- Gain experience diagnosing MCMC chains visually, with Rhat, and with ESS.

# Exercise

For each of the following sets of saved MCMC chains, answer the following:

1. Do these samples appear consistent with convergence?
2. What helped you come to that conclusion?

If inconsistent with convergence, continue:

3. What might have caused this scenario?
4. Can the MCMC sampling likely be improved without adjusting the model? How?

```{r mcmc-chains1}
samples <- readRDS(here::here("data/mcmc1.rds"))
samples_matrix <- do.call(cbind, samples)

bayesplot::mcmc_trace(samples)
rstan::Rhat(samples_matrix)
rstan::ess_bulk(samples_matrix)
rstan::ess_tail(samples_matrix)
```

1. No <!-- exercise -->
2. Rhat a bit high, ESS a bit low, strong autocorrelation apparent in chains <!-- exercise -->
3. Chains not run for long enough <!-- exercise -->
4. Possibly just running for longer; no thinning needed! <!-- exercise -->

```{r mcmc-chains2}
samples <- readRDS(here::here("data/mcmc2.rds"))
samples_matrix <- do.call(cbind, samples)

bayesplot::mcmc_trace(samples)
rstan::Rhat(samples_matrix)
rstan::ess_bulk(samples_matrix)
rstan::ess_tail(samples_matrix)
```

1. No <!-- exercise -->
2. Chains drifting, Rhat high, ESS low <!-- exercise -->
3. Possibly not enough warmup and starting points not dispersed enough <!-- exercise -->
4. Longer warmup? <!-- exercise -->

```{r mcmc-chains3}
samples <- readRDS(here::here("data/mcmc3.rds"))
samples_matrix <- do.call(cbind, samples)

bayesplot::mcmc_trace(samples)
bayesplot::mcmc_dens_overlay(samples)
rstan::Rhat(samples_matrix)
rstan::ess_bulk(samples_matrix)
rstan::ess_tail(samples_matrix)
```

1. No <!-- exercise -->
2. One chain has larger variance <!-- exercise -->
3. Some chains may be getting stuck and not exploring posterior <!-- exercise -->
4. May be harder. Tighter priors, re-parameterized model, simplified model? <!-- exercise -->

```{r mcmc-chains4}
samples <- readRDS(here::here("data/mcmc4.rds"))
samples_matrix <- do.call(cbind, samples)

bayesplot::mcmc_trace(samples)
bayesplot::mcmc_dens_overlay(samples)
rstan::Rhat(samples_matrix)
rstan::ess_bulk(samples_matrix)
rstan::ess_tail(samples_matrix)
```

1. Yes <!-- exercise -->
2. ESS > 400, Rhat < 1.01, chains look good <!-- exercise -->
3. NA <!-- exercise -->
4. NA <!-- exercise -->

```{r mcmc-chains5}
samples <- readRDS(here::here("data/mcmc5.rds"))
samples_matrix <- do.call(cbind, samples)

bayesplot::mcmc_trace(samples)
rstan::Rhat(samples_matrix)
rstan::ess_bulk(samples_matrix)
rstan::ess_tail(samples_matrix)
bayesplot::mcmc_dens_overlay(samples)
```

1. No <!-- exercise -->
2. Chains have different means <!-- exercise -->
3. Might still reflect dispersed starting points without enough warmup? Model might be weakly identified? <!-- exercise -->
4. Longer warmup? Simplify model? More data? <!-- exercise -->

```{r mcmc-chains6}
samples <- readRDS(here::here("data/mcmc6.rds"))
samples_matrix <- do.call(cbind, samples)

bayesplot::mcmc_trace(samples)
rstan::Rhat(samples_matrix)
rstan::ess_bulk(samples_matrix)
rstan::ess_tail(samples_matrix)
```

1. No <!-- exercise -->
2. Chain behaviour at beginning looks different, autocorrelation may be high <!-- exercise -->
3. Might have forgot to remove warmup or not run warmup for long enough? <!-- exercise -->
4. Discard warmup, warmup long enough, ... <!-- exercise -->
