---
title: "Priors in brms"
output:
  html_document:
    toc: true
    toc_float: true
---

# Goals:

- Learn how to find what priors are needed in a given brms model and what the defaults are.
- Learn how to specify priors in brms.
- Learn how to see the influence of priors on the posterior.
- Learn how to do a prior predictive check.
- Learn how brms treats intercepts.

# Exercise:

```{r}
library(ggplot2)
library(dplyr)
theme_set(theme_light())
library(brms)
dir.create("cache", showWarnings = FALSE)
options(brms.file_refit = "on_change") # re-fit cached models if changes
```

Let's create some fake data:

```{r}
x <- 1:20
n <- length(x)
a <- 0.2
b <- 0.3
sigma <- 1
set.seed(2141)
y <- a + b * x + sigma * rnorm(n)
fake <- data.frame(x, y)
```

```{r}
ggplot(fake, aes(x, y)) + geom_point()
```

If we wanted to fit a simple linear regression of y on x, we could find the default priors for the parameters with:

```{r}
brms::default_prior(
  y ~ x, data = fake
)
```

If we fit out model with the default priors:

```{r}
mod1 <- brm(y ~ x, data = fake, family = gaussian())
```

We could find out what priors we used with:

```{r}
get_prior(mod1)
```

We should instead specify our own priors. We can do this with the `prior` argument.

Note that by default, brms parameterizes the intercept as the intercept when the predictors are at their mean:

```{r, eval=FALSE}
mod1 <- brm(y ~ x, data = fake, family = gaussian(),
  file = "cache/priors1",
  prior = c(
    set_prior("normal(0, 1)", class = "Intercept"),
    set_prior("normal(0, 1)", class = "b"),
    set_prior("student_t(3, 0, 3)", class = "sigma")
  ),
  sample_prior = "yes"
)
```

If instead we wanted to parameterize the intercept as the value of y when our predictors are 0, we would need to use the `0 + Intercept` syntax:

```{r}
mod2 <- brm(y ~ 0 + Intercept + x, data = fake, family = gaussian(),
  file = "cache/priors2",
  prior = c(
    set_prior("normal(0, 1)", class = "b", coef = "Intercept"),
    set_prior("normal(0, 1)", class = "b", coef = "x"),
    set_prior("student_t(3, 0, 3)", class = "sigma")
  ),
  sample_prior = "yes"
)
```

We can look at the what's happening in the Stan code:

```{r}
stancode(mod1)
stancode(mod2)
```

```{r}
sims1 <- as_draws_df(mod1)
mean(sims1$b_Intercept)
mean(sims1$Intercept)
mean(fake$y)
```

```{r}
sims2 <- as_draws_df(mod2)
mean(sims2$b_Intercept)
mean(fake$y)
```

```{r}
p2 <- brms::prior_draws(mod2) # sample_prior = "yes"
hist(p2$b_x, breaks = 50)
hist(p2$sigma, breaks = 50)
```

```{r}
mod_prior_only <- brm(
  y ~ 0 + Intercept + x, data = fake, family = gaussian(),
  file = "cache/priors3",
  prior = c(
    set_prior("normal(0, 5)", class = "b", coef = "Intercept"),
    set_prior("normal(0, 1)", class = "b", coef = "x"),
    set_prior("student_t(3, 0, 3)", class = "sigma")
  ),
  sample_prior = "only", #<
  seed = 2028, iter = 100, chains = 1
)
```

```{r}
prior_pushforward <- brms::posterior_epred(mod_prior_only)
prior_predictive <- brms::posterior_predict(mod_prior_only)
```

```{r}
fake$pf1 <- prior_pushforward[1,]
fake$pf2 <- prior_pushforward[2,]
fake$pf3 <- prior_pushforward[3,]
ggplot(fake, aes(x, pf1)) + geom_point()
ggplot(fake, aes(x, pf2)) + geom_point()
ggplot(fake, aes(x, pf3)) + geom_point()
```

```{r}
fake$pp1 <- prior_predictive[2,]
fake$pp2 <- prior_predictive[3,]
fake$pp3 <- prior_predictive[4,]
ggplot(fake, aes(x, pp1)) + geom_point()
ggplot(fake, aes(x, pp2)) + geom_point()
ggplot(fake, aes(x, pp3)) + geom_point()
```

```{r}
mod_prior_only_wide <- brm(
  y ~ 0 + Intercept + x, data = fake, family = gaussian(),
  file = "cache/priors4",
  prior = c(
    set_prior("normal(0, 100)", class = "b", coef = "Intercept"),
    set_prior("normal(0, 100)", class = "b", coef = "x"),
    set_prior("student_t(3, 0, 100)", class = "sigma")
  ),
  sample_prior = "only",
  seed = 2028, iter = 100, chains = 1
)
```

```{r}
prior_predictive <- brms::posterior_predict(mod_prior_only_wide)
fake$pp1 <- prior_predictive[1,]
fake$pp2 <- prior_predictive[18,]
fake$pp3 <- prior_predictive[42,]
ggplot(fake, aes(x, pp1)) + geom_point()
ggplot(fake, aes(x, pp2)) + geom_point()
ggplot(fake, aes(x, pp3)) + geom_point()
```
